Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-11 _paper.rmd
updated penguin case study but it will need more work

---
title: "Exploring Local Explanations of Non-linear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
  #- name: "Nicholas Spyrison"
  #   address: |
  #     | ORCiD: 0000-0002-8417-0212
  #   email: \email{nicholas.spyrison@monash.edu}
  #   orcid: 0000-0002-8417-0212
  #- name: "Dianne Cook"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 000-0002-3813-7155
  #   orcid: 000-0002-3813-7155
  #- name: "Kimbal Marriott"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 0000-0002-9813-0377
  #   orcid: 0000-0002-9813-0377
abstract: |
 Artificial Intelligence (AI) has seen a revitalization in recent years with the prevalence of increasingly hard-to-interpret black-box models. The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear variable importance in the vicinity of each observation. These are simply linear projections, and as such can be further explored interactively to better understand the interaction between variables used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response variables. The methods are implemented in the __R__ package __cheem__, available on CRAN.
#documentclass: jmlr
#classoption: article
output:
  ## pdf_document: requires \ref{} 
  #pdf_document:
  ## bookdown::pdf_document2: allows \@ref()
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    number_sections: yes
    fig_caption: yes
    #includes:
    #  in_header: template/jmlr2e.sty
bibliography: cheem_paper.bib
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
  - \usepackage{setspace} ## for title page spacing
  - \usepackage{hyperref} ## for all sorts of linking
---

@@

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "The cheem view for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified observations. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear variable importances that best describe the observation's prediction. We probe the support of variable contributions by selecting a variable and use the radial tour to vary its coefficient."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

```{r regressioncase, echo=F, out.width = "100%", fig.cap="Overview of the cheem view for tasks. The global view can be colored on the correlation between the projection of the data generated by the local explanation and the observed response. In the radial tour, the horizontal values are the same as the classification case; the projection through the basis. The vertical position is now mapped to the observed y and residuals."}
knitr::include_graphics("./figures/app_regression.PNG")
```

@@

The Palmer penguins data [@gorman_ecological_2014; @horst_palmerpenguins_2020] was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was provided publicly as a substitute for the over-used iris data, and is quite similar in form. There are 330 observations and we will use the four physical measurements, `bill_length_mm` (`b_l`), `bill_depth_mm` (`b_d`), `flipper_length_mm` (`f_l`), `body_mass_g` (`wgt`), for this illustration. A random forest model was fitted with species as the response variable.

(ref:casepenguins) Species classification of Palmer penguin data. We select a adelie penguin that is mislabeled as an chinstrap. Starting from the attribution projection we varying the contribution to `bill length`. This explanation is not supported sense by the time there is moderate contribution. The current basis is fairly close to the regularly attribution to other chinstrap penguins aside from the sign on flipper length being flipped. It feels odd that SHAP has identified a profile where chinstrap and adelie penguins are hard to distinguish when there are separating orientations that should be found within 125 trees. The .mp4 animation of this tour can be found at [shorturl.at/hFHNS](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins)"}
knitr::include_graphics("./figures/case_penguins.png")
```

Figure \@ref(fig:casepenguins) shows plots from the cheem viewer for exploring the random forest model on the penguins data. Plot (a) shows the global view, and plot (b) shows several plots generated with the radial tour.

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified observations. A misclassified point (represented by the asterisk in the global view, and as a solid vertical line in the tour view) is contrasted with a correctly classified point (represented by an $x$ and vertical dashed line).

The radial tour starts from the attribution projection of the misclassified observation (left plot). The important variables identifed by SHAP are `b_d` and `wgt` in contrast to `f_l`.

varies the contribution of bill length (b_l) as this variable differs most from the contribution of the comparison observation. Downplaying the contribution of bill length is crucial to the linear explanation of this observation being misclassified.

>>>>

---
title: "Exploring Local Explanations of Non-linear Models Using Animated Linear Projections"
shorttitle: Cheem
keywords:
- R
- XAI
- explanable artificial intelligence
- black-box models
- local explanations
- grand tour
- manual tour
- radial tour
author:
  #- name: "Nicholas Spyrison"
  #   address: |
  #     | ORCiD: 0000-0002-8417-0212
  #   email: \email{nicholas.spyrison@monash.edu}
  #   orcid: 0000-0002-8417-0212
  #- name: "Dianne Cook"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 000-0002-3813-7155
  #   orcid: 000-0002-3813-7155
  #- name: "Kimbal Marriott"
  #   affiliation: Monash University
  #   address: |
  #     | ORCiD: 0000-0002-9813-0377
  #   orcid: 0000-0002-9813-0377
abstract: |
 The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear variable importance in the vicinity of each observation. These can be considered to be linear projections, and as such can be further explored interactively to better understand the interaction between variables used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response variables. The methods are implemented in the __R__ package __cheem__, available on CRAN.
#documentclass: jmlr
#classoption: article
output:
  ## pdf_document: requires \ref{} 
  #pdf_document:
  ## bookdown::pdf_document2: allows \@ref()
  bookdown::pdf_document2:
    keep_tex: true
    toc: false
    number_sections: yes
    fig_caption: yes
    #includes:
    #  in_header: template/jmlr2e.sty
bibliography: cheem_paper.bib
editor_options:
  chunk_output_type: console
urlcolor: blue
linkcolor: red
header-includes:
  - \usepackage{setspace} ## for title page spacing
  - \usepackage{hyperref} ## for all sorts of linking
---

@@

```{r classificationcase, echo=F, out.width = "100%", fig.align="center", fig.cap = "Overview of the cheem viewer for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified observations. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear variable importances that best describe the observation's prediction. We probe the support of variable contributions by selecting a variable and use the radial tour to vary its coefficient."}
knitr::include_graphics("./figures/app_classification.PNG")
```

@@

```{r regressioncase, echo=F, out.width = "100%", fig.cap="Overview of the cheem viewer for regression tasks. The global view can be colored on the correlation between the projection of the data generated by the local explanation and the observed response. In the radial tour, the horizontal values are the same as the classification case; the projection through the basis. The vertical position is now mapped to the observed y and residuals."}
knitr::include_graphics("./figures/app_regression.PNG")
```

@@

The Palmer penguins data [@gorman_ecological_2014; @horst_palmerpenguins_2020] was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to be a substitute for the overly-used iris data, and is quite similar in form. There are 330 observations and we will use the four physical measurements, `bill_length_mm` (`b_l`), `bill_depth_mm` (`b_d`), `flipper_length_mm` (`f_l`), `body_mass_g` (`wgt`), for this illustration. A random forest model was fitted with species as the response variable.

(ref:casepenguins) Examining the SHAP values for a random forest model classifying Palmer penguin species. The observation of interest is an Adelie (green) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). XXX needs fixing Starting from the attribution projection the contribution of `b_l` is varied. This explanation is not supported sense by the time there is moderate contribution. The current basis is fairly close to the regularly attribution to other chinstrap penguins aside from the sign on flipper length being flipped. It feels odd that SHAP has identified a profile where chinstrap and adelie penguins are hard to distinguish when there are separating orientations that should be found within 125 trees. The .mp4 animation of this tour can be found at [shorturl.at/hFHNS](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4)
```{r casepenguins, echo=F, out.width="100%", fig.cap = "(ref:casepenguins)"}
knitr::include_graphics("./figures/case_penguins.png")
```

Figure \@ref(fig:casepenguins) shows plots from the cheem viewer for exploring the random forest model on the penguins data. Plot (a) shows the global view, and plot (b) shows several 1D projections generated with the radial tour. Penguin 124, an Adelie (green), is the observation of interest because it has been misclassified as a Chinstrap (orange).

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified observations. A misclassified point (represented by the asterisk in the global view, and as a dashed vertical line in the tour view) is contrasted with a correctly classified point (represented by an $\times$ and vertical dotted line).

The radial tour starts from the attribution projection of the misclassified observation (left plot). The important variables identifed by SHAP in the (wrong) prediction for this observation is mostly `b_d` with smaller contributions of `f_l` and `wgt`. This projection is a view where the Gentoo (purple) penguins are most distinguished from Adelie and Chinstrap. For this observation we would like to know why an Adelie has been wrongly labelled a Chinstrap, so this combination of variables is not particularly useful. To explore this we use the radial tour to vary the contribution of bill length (`b_l`). The more `b_l` contributes to the the projection the more this penguin looks like a Chinstrap, that it is mislabelled as. This suggests that `b_l` should have been an important variable for explaining the (wrong) prediction.

### 2022-01-11 _paper.tex
updated penguin case study but it will need more work

\begin{document}
\maketitle
\begin{abstract}
Artificial Intelligence (AI) has seen a revitalization in recent years with the prevalence of increasingly hard-to-interpret black-box models. The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear variable importance in the vicinity of each observation. These are simply linear projections, and as such can be further explored interactively to better understand the interaction between variables used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response variables. The methods are implemented in the \textbf{R} package \textbf{cheem}, available on CRAN.
\end{abstract}

@@

\caption{The cheem view for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified observations. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear variable importances that best describe the observation's prediction. We probe the support of variable contributions by selecting a variable and use the radial tour to vary its coefficient.}\label{fig:classificationcase}
\end{figure}

@@

\caption{Overview of the cheem view for tasks. The global view can be colored on the correlation between the projection of the data generated by the local explanation and the observed response. In the radial tour, the horizontal values are the same as the classification case; the projection through the basis. The vertical position is now mapped to the observed y and residuals.}\label{fig:regressioncase}
\end{figure}

@@

The Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was provided publicly as a substitute for the over-used iris data, and is quite similar in form. There are 330 observations and we will use the four physical measurements, \texttt{bill\_length\_mm} (\texttt{b\_l}), \texttt{bill\_depth\_mm} (\texttt{b\_d}), \texttt{flipper\_length\_mm} (\texttt{f\_l}), \texttt{body\_mass\_g} (\texttt{wgt}), for this illustration. A random forest model was fitted with species as the response variable.

@@

\caption{Species classification of Palmer penguin data. We select a chinstrap penguin that is mislabeled as an adelie. By varying the contribution to bill length, we observe the explanation does not hold when bill length has a significant contribution. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}}\label{fig:casepenguins}
\end{figure}

Figure \ref{fig:casepenguins} shows plots from the cheem viewer for exploring the random forest model on the penguins data. Plot (a) shows the global view, and plot (b) shows several plots generated with the radial tour.

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified observations. A misclassified point (represented by the asterisk in the global view, and as a solid vertical line in the tour view) is contrasted with a correctly classified point (represented by an \(x\) and vertical dashed line).

The radial tour starts from the attribution projection of the misclassified observation (left plot). The important variables identifed by SHAP are \texttt{b\_d} and \texttt{wgt} in contrast to \texttt{f\_l}.

varies the contribution of bill length (b\_l) as this variable differs most from the contribution of the comparison observation. Downplaying the contribution of bill length is crucial to the linear explanation of this observation being misclassified.

\hypertarget{chocolates-milkdark-chocolate-classification}{%
\subsection{2) Chocolates, milk/dark chocolate classification}\label{chocolates-milkdark-chocolate-classification}}

The chocolates dataset consists of 88 observations of 10 nutritional measurements from their labels. Each of which was labeled as being either milk or dark chocolates. We can see if a manufacturer accurately portrays the chocolate with this data. We are curious to see if chocolates that nutritionally look like milk chocolates are labeled as dark chocolates, which may hold a higher market value. We should note that not all chocolates consist wholly of chocolate. The addition of other ingredients will decrease the predictive power of the model nutritional explanatory variable. A random forest model is fit classifying the type of chocolate. We selected a chocolate labeled dark, through predicted to be milk chocolate compared with a chocolate labeled 85\% cocoa.

@@

>>>>

\title{Exploring Local Explanations of Non-linear Models Using Animated Linear Projections}
\author{}
\date{\vspace{-2.5em}}

\begin{document}
\maketitle
\begin{abstract}
The increased predictive power comes at the cost of interpretability, which has led to the emergence of eXplainable AI (XAI). XAI attempts to shed light on how models are using predictors, to arrive at a prediction, with a point estimate of the linear variable importance in the vicinity of each observation. These can be considered to be linear projections, and as such can be further explored interactively to better understand the interaction between variables used to make predictions, across the predictive model surface. Here we describe interactive linear interpolation used for exploration at any observation, and illustrate with examples with categorical (penguin species, chocolate types) and quantitative (football salaries, house prices) response variables. The methods are implemented in the \textbf{R} package \textbf{cheem}, available on CRAN.
\end{abstract}

@@

\caption{Overview of the cheem viewer for classification tasks. Plots are colored on predicted class, and red circles indicate misclassified observations. Model visual is essentially a representation of the confusion matrix of the model. The radial tour is a 1D projection starting the attribution projection. The first frame is the linear variable importances that best describe the observation's prediction. We probe the support of variable contributions by selecting a variable and use the radial tour to vary its coefficient.}\label{fig:classificationcase}
\end{figure}

@@

\caption{Overview of the cheem viewer for regression tasks. The global view can be colored on the correlation between the projection of the data generated by the local explanation and the observed response. In the radial tour, the horizontal values are the same as the classification case; the projection through the basis. The vertical position is now mapped to the observed y and residuals.}\label{fig:regressioncase}
\end{figure}

@@

The Palmer penguins data (Gorman, Williams, and Fraser 2014; Horst, Hill, and Gorman 2020) was collected on three species of penguins foraging near Palmer Station, Antarctica. The data was publicly available to be a substitute for the overly-used iris data, and is quite similar in form. There are 330 observations and we will use the four physical measurements, \texttt{bill\_length\_mm} (\texttt{b\_l}), \texttt{bill\_depth\_mm} (\texttt{b\_d}), \texttt{flipper\_length\_mm} (\texttt{f\_l}), \texttt{body\_mass\_g} (\texttt{wgt}), for this illustration. A random forest model was fitted with species as the response variable.

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The observation of interest is an Adelie (green) penguin that is misclassified as a Chinstrap (orange), marked as an asterisk in (a), and the dashed vertical line in (b). XXX needs fixing Starting from the attribution projection the contribution of \texttt{b\_l} is varied. This explanation is not supported sense by the time there is moderate contribution. The current basis is fairly close to the regularly attribution to other chinstrap penguins aside from the sign on flipper length being flipped. It feels odd that SHAP has identified a profile where chinstrap and adelie penguins are hard to distinguish when there are separating orientations that should be found within 125 trees. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}}\label{fig:casepenguins}
\end{figure}

Figure \ref{fig:casepenguins} shows plots from the cheem viewer for exploring the random forest model on the penguins data. Plot (a) shows the global view, and plot (b) shows several 1D projections generated with the radial tour. Penguin 124, an Adelie (green), is the observation of interest because it has been misclassified as a Chinstrap (orange).

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified observations. A misclassified point (represented by the asterisk in the global view, and as a dashed vertical line in the tour view) is contrasted with a correctly classified point (represented by an \(\times\) and vertical dotted line).

The radial tour starts from the attribution projection of the misclassified observation (left plot). The important variables identifed by SHAP in the (wrong) prediction for this observation is mostly \texttt{b\_d} with smaller contributions of \texttt{f\_l} and \texttt{wgt}. This projection is a view where the Gentoo (purple) penguins are most distinguished from Adelie and Chinstrap. For this observation we would like to know why an Adelie has been wrongly labelled a Chinstrap, so this combination of variables is not particularly useful. To explore this we use the radial tour to vary the contribution of bill length (\texttt{b\_l}). The more \texttt{b\_l} contributes to the the projection the more this penguin looks like a Chinstrap, that it is mislabelled as. This suggests that \texttt{b\_l} should have been an important variable for explaining the (wrong) prediction.

\hypertarget{chocolates-milkdark-chocolate-classification}{%
\subsection{Chocolates, milk/dark chocolate classification}\label{chocolates-milkdark-chocolate-classification}}

The chocolates dataset consists of 88 observations of ten nutritional measurements from their labels. Each of which was labeled as being either milk or dark chocolates. We can see if a manufacturer accurately portrays the chocolate with this data. We are curious to see if chocolates that nutritionally look like milk chocolates are labeled as dark chocolates, which may hold a higher market value. We should note that not all chocolates consist wholly of chocolate. The addition of other ingredients will decrease the predictive power of the model nutritional explanatory variable. A random forest model is fit classifying the type of chocolate. We selected a chocolate labeled dark, through predicted to be milk chocolate compared with a chocolate labeled 85\% cocoa.

@@

\leavevmode\vadjust pre{\hypertarget{ref-gabriel_biplot_1971}{}}%
Gabriel, Karl Ruben. 1971. {``The Biplot Graphic Display of Matrices with Application to Principal Component Analysis.''} \emph{Biometrika} 58 (3): 453--67.
